quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-06-05 paper-quollr.Rmd
Need to work on the best model

<!--add intro to prism data-->

To illustrate the method, we use $5\text{-}D$ simulated data, which we call the "triangular_3d_data". This dataset is generated using the `tri_3d` function from the `cardinalR` package. The dataset generation starts with the initialization of a starting point $p_0 = (x_0, y_0, z_0)$ randomly selected within the $3\text{-}D$ space using a uniform distribution. Four fixed corner points of a tetrahedron are defined as $c_1 = (0, 0, 0)$, $c_2 = (1, 0, 0)$, $c_3 = (0.5, 1, 0)$, and $c_4 = (0.5, 0.5, 1)$. For each point $p_i$ where $i = 1, 2, \ldots, n$, one of the corner points $c_j$ (where $j \in \{1, 2, 3, 4\}$) is selected randomly, and the new point $p_i$ is computed as the midpoint between the current point $p_{i-1}$ and the selected corner point $c_j$: $p_i = \frac{p_{i-1} + c_j}{2}$. This iterative process ensures that each new point moves closer to one of the corners, creating a fractal-like triangular distribution of points. The coordinates of each generated point $p_i$ form the variables $X_1, X_2, X_3$. The remaining variables $X_4, X_5$ are all uniform error, with small variance. We would consider $T=(X_1, X_2, X_3)$ to be the true model.

<!-- how the data are generated and from which package I got the data-->

@@

<!--added, I selected tSNE with perplexity = 35 to explore and -->

```{r}
sc_ltr_pos <- c(0.08, 0.96)

tSNE_fit <- triangular_3d_data |>
  select(-ID) |>
  Rtsne(perplexity = round(sqrt(NROW(triangular_3d_data))))

@@

nldr_prism1 <- tsne_prism_scaled |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5)

nldr_prism1
```

<!--select based on looking at b_non_empty vs log(MSE)-->
<!--select a choice for b_non_empty-->

<!--with prism data-->
<!-- my plan is to discuss which the appropriate representation from different parameter choices from tSNE-->
<!--Why tSNE? thinking about the usability-->

<!--To generate errors for different total number of bins-->
```{r}
## To initialize number of bins along the x-axis
bin1_vec_prism <- 2:25

@@

error_prism <- data.frame(matrix(nrow = 0, ncol = 0))

@@

  ## Compute error
  error_df <- glance(
    df_bin_centroids = df_bin_centroids_prism,
    df_bin = df_bin_prism,
    training_data = triangular_3d_data,
    newdata = NULL,
    type_NLDR = "tSNE",
    col_start = "x") |>
    mutate(bin1 = xbins,
           bin2 = bin2,
           b = bin1 * bin2,
           b_non_empty = NROW(df_bin_centroids_prism))

  error_prism <- bind_rows(error_prism, error_df)

@@

mse_prism_b <- ggplot(error_prism, 
                     aes(x = b, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 420, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  ylab("log(MSE)") +
  xlab("b") +
  theme(axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7, angle = 90))

mse_prism_b
```

```{r}
## Compute hexbin parameters
num_bins_x_prism <- 20
lim1 <- prism_scaled_obj$lim1
lim2 <- prism_scaled_obj$lim2
r2_prism <- diff(lim2)/diff(lim1)

prism_model <- fit_highd_model(
  training_data = triangular_3d_data,
  emb_df = tsne_prism_scaled,
  bin1 = num_bins_x_prism,
  r2 = r2_prism,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "x"
)

@@

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_prism <- hex_binning(
  data = tsne_prism_scaled,
  bin1 = num_bins_x_prism,
  r2 = r2_prism)

>>>>

To illustrate the algorithm, we use $5\text{-}D$ simulated data, which we call the "triangular_3d_data". This dataset is generated using the `tri_3d` function from the `cardinalR` package. The dataset generation starts with the initialization of a starting point $p_0 = (x_0, y_0, z_0)$ randomly selected within the $3\text{-}D$ space using a uniform distribution. Four fixed corner points of a tetrahedron are defined as $c_1 = (0, 0, 0)$, $c_2 = (1, 0, 0)$, $c_3 = (0.5, 1, 0)$, and $c_4 = (0.5, 0.5, 1)$. For each point $p_i$ where $i = 1, 2, \ldots, n$, one of the corner points $c_j$ (where $j \in \{1, 2, 3, 4\}$) is selected randomly, and the new point $p_i$ is computed as the midpoint between the current point $p_{i-1}$ and the selected corner point $c_j$: $p_i = \frac{p_{i-1} + c_j}{2}$. This iterative process ensures that each new point moves closer to one of the corners, creating a fractal-like triangular distribution of points. The coordinates of each generated point $p_i$ form the variables $X_1, X_2, X_3$. The remaining variables $X_4, X_5$ are all uniform error, with small variance. We would consider $T=(X_1, X_2, X_3)$ to be the true model.

@@

```{r}
sc_ltr_pos <- c(0.08, 0.96)
```

<!--tSNE with perplexity = 6-->
```{r}
tSNE_fit1 <- triangular_3d_data |>
  select(-ID) |>
  Rtsne(perplexity = 6)

tsne_prism1 <- tSNE_fit1$Y |>
  as_tibble() |>
  mutate(ID = row_number())

colnames(tsne_prism1) <- c("tSNE1", "tSNE2", "ID")

prism_scaled_obj_tsne1 <- gen_scaled_data(
  data = tsne_prism1)
tsne_prism_scaled1 <- prism_scaled_obj_tsne1$scaled_nldr

nldr_prism1 <- tsne_prism_scaled1 |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

<!--tSNE with perplexity = 35-->
```{r}
tSNE_fit2 <- triangular_3d_data |>
  select(-ID) |>
  Rtsne(perplexity = round(sqrt(NROW(triangular_3d_data))))

tsne_prism2 <- tSNE_fit2$Y |>
  as_tibble() |>
  mutate(ID = row_number())

colnames(tsne_prism2) <- c("tSNE1", "tSNE2", "ID")

prism_scaled_obj_tsne2 <- gen_scaled_data(
  data = tsne_prism2)
tsne_prism_scaled2 <- prism_scaled_obj_tsne2$scaled_nldr

nldr_prism2 <- tsne_prism_scaled2 |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

<!--tSNE with perplexity = 58-->
```{r}
tSNE_fit3 <- triangular_3d_data |>
  select(-ID) |>
  Rtsne(perplexity = 58)

tsne_prism3 <- tSNE_fit3$Y |>
  as_tibble() |>
  mutate(ID = row_number())

colnames(tsne_prism3) <- c("tSNE1", "tSNE2", "ID")

prism_scaled_obj_tsne3 <- gen_scaled_data(
  data = tsne_prism3)
tsne_prism_scaled3 <- prism_scaled_obj_tsne3$scaled_nldr

nldr_prism3 <- tsne_prism_scaled3 |> 
  ggplot(aes(x = tSNE1, y = tSNE2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

<!--UMAP with n_neighbors = 15 and min_dist = 0.3-->
```{r}
umap_prism1 <- triangular_3d_data |>
  select(-ID) |>
  umap(n_neighbors = 15, n_components =  2, 
       metric = "euclidean", min_dist = 0.3, init = "spca")

umap_prism1 <- umap_prism1 |>
  as_tibble()

names(umap_prism1) <- c("UMAP1", "UMAP2")

umap_prism1 <- umap_prism1 |>
  mutate(ID = 1:NROW(umap_prism1))

prism_scaled_obj_umap1 <- gen_scaled_data(
  data = umap_prism1)
umap_prism_scaled1 <- prism_scaled_obj_umap1$scaled_nldr

nldr_prism4 <- umap_prism_scaled1 |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

<!--UMAP with n_neighbors = 35 and min_dist = 0.5-->
```{r}
umap_prism2 <- triangular_3d_data |>
  select(-ID) |>
  umap(n_neighbors = 35, n_components =  2, 
       metric = "euclidean", min_dist = 0.5, init = "spca")

umap_prism2 <- umap_prism2 |>
  as_tibble()

names(umap_prism2) <- c("UMAP1", "UMAP2")

umap_prism2 <- umap_prism2 |>
  mutate(ID = 1:NROW(umap_prism2))

prism_scaled_obj_umap2 <- gen_scaled_data(
  data = umap_prism2)
umap_prism_scaled2 <- prism_scaled_obj_umap2$scaled_nldr

nldr_prism5 <- umap_prism_scaled2 |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

<!--UMAP with n_neighbors = 69 and min_dist = 0.6-->
```{r}
umap_prism3 <- triangular_3d_data |>
  select(-ID) |>
  umap(n_neighbors = 69, n_components =  2, 
       metric = "euclidean", min_dist = 0.6, init = "spca")

umap_prism3 <- umap_prism3 |>
  as_tibble()

names(umap_prism3) <- c("UMAP1", "UMAP2")

umap_prism3 <- umap_prism3 |>
  mutate(ID = 1:NROW(umap_prism3))

prism_scaled_obj_umap3 <- gen_scaled_data(
  data = umap_prism3)
umap_prism_scaled3 <- prism_scaled_obj_umap3$scaled_nldr

nldr_prism6 <- umap_prism_scaled3 |> 
  ggplot(aes(x = UMAP1, y = UMAP2)) + 
  geom_point(alpha=0.5, colour="#000000", size = 0.5) 
```

```{r}
nldr_prism1 + nldr_prism2 + nldr_prism3 + 
  nldr_prism4 + nldr_prism5 + nldr_prism6 +
  plot_layout(ncol=3)
```

<!--need to add details regarding which differences can I see with different layouts-->

<!--select based on looking at b_non_empty vs log(Error)-->
<!--select a choice for b_non_empty-->

<!--compute error with tSNE and different perplexity values-->
```{r}
error_prism_tsne <- data.frame(matrix(nrow = 0, ncol = 0))

## To initialize number of bins along the x-axis
bin1_vec_prism <- 2:25

perplexity_vec <- c(6, 35, 58)

for (perplexity_val in perplexity_vec) {

  tSNE_fit <- triangular_3d_data |>
    select(-ID) |>
    Rtsne(perplexity = perplexity_val)

@@

@@

@@

    ## Compute error
    error_df <- glance(
      df_bin_centroids = df_bin_centroids_prism,
      df_bin = df_bin_prism,
      training_data = triangular_3d_data,
      newdata = NULL,
      type_NLDR = "tSNE",
      col_start = "x") |>
      mutate(bin1 = xbins,
             bin2 = bin2,
             b = bin1 * bin2,
             b_non_empty = NROW(df_bin_centroids_prism),
             type = paste0("perplexity: ", perplexity_val))

    error_prism_tsne <- bind_rows(error_prism_tsne, error_df)

@@

}
```

<!--compute error with UMAP and different parameter values-->
```{r}
error_prism_umap <- data.frame(matrix(nrow = 0, ncol = 0))

## To initialize number of bins along the x-axis
bin1_vec_prism <- 2:25

param_list <- list(param1 = c(15, 0.3), param2 = c(35, 0.5), param3 = c(69, 0.6))

for (i in 1:length(param_list)) {

  n_neighbors_val <- param_list[[i]][1]
  min_dist_val <- param_list[[i]][2]

  umap_prism <- triangular_3d_data |>
    select(-ID) |>
    umap(n_neighbors = n_neighbors_val, n_components =  2, 
         metric = "euclidean", min_dist = min_dist_val, init = "spca")

  umap_prism <- umap_prism |>
    as_tibble()

  names(umap_prism) <- c("UMAP1", "UMAP2")

  umap_prism <- umap_prism |>
    mutate(ID = 1:NROW(umap_prism))

  prism_scaled_obj <- gen_scaled_data(
    data = umap_prism)
  umap_prism_scaled <- prism_scaled_obj$scaled_nldr

    prism_model <- fit_highd_model(
      training_data = triangular_3d_data,
      emb_df = umap_prism_scaled,
      bin1 = xbins,
      r2 = r2_prism,
      is_bin_centroid = TRUE,
      is_rm_lwd_hex = TRUE,
      col_start_highd = "x"
    )

    ## Compute error
    error_df <- glance(
      df_bin_centroids = df_bin_centroids_prism,
      df_bin = df_bin_prism,
      training_data = triangular_3d_data,
      newdata = NULL,
      type_NLDR = "UMAP",
      col_start = "x") |>
      mutate(bin1 = xbins,
             bin2 = bin2,
             b = bin1 * bin2,
             b_non_empty = NROW(df_bin_centroids_prism),
             type = paste0("n_neighbors: ", n_neighbors_val, 
                           " min_dist: ", min_dist_val))

    error_prism_umap <- bind_rows(error_prism_umap, error_df)

}
```

```{r}
error_prism <- bind_rows(error_prism_tsne, error_prism_umap) 
error_prism <- error_prism |>
  filter(b <= 625)
```

```{r}
error_plot_prism <- ggplot(error_prism, 
                           aes(x = b_non_empty, 
                               y = log(Error), 
                               group = type, 
                               colour = type)) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 112, linetype="solid",
             color = "black", linewidth=0.8, alpha = 0.5) +
  scale_color_manual(values=c('#e41a1c','#377eb8','#4daf4a','#984ea3',
                              '#ff7f00','#a6cee3','#a65628','#f781bf')) +
  ylab("log(Error)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        legend.position = "bottom")

error_plot_prism
```

<!--selected tSNE perplexity: 35-->

```{r}
## Compute hexbin parameters
num_bins_x_prism <- 21
lim1 <- prism_scaled_obj_tsne2$lim1
lim2 <- prism_scaled_obj_tsne2$lim2
r2_prism <- diff(lim2)/diff(lim1)

prism_model <- fit_highd_model(
  training_data = triangular_3d_data,
  emb_df = tsne_prism_scaled2,
  bin1 = num_bins_x_prism,
  r2 = r2_prism,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = TRUE,
  col_start_highd = "x"
)

@@

```{r}
## Hexagonal binning to have regular hexagons
hb_obj_prism <- hex_binning(
  data = tsne_prism_scaled2,
  bin1 = num_bins_x_prism,
  r2 = r2_prism)
